%---------------------
\section{Introduction}
\label{sec:introduction}
%---------------------
Healthcare expenditure in the US is growing concern. The total annual healthcare expenditure in 2012 was \$2.8 trillion, the prescription drug spending accounting for \$260.8 billion \cite{healthaffairs}. As the percentage of GDP (17.9), it is the highest among all nations. The healthcare cost per capita in 2010 was \$8,233, which was one-fifth of the personal income per capita (\$42,693). The health expenditure per capita was about \$3,000 more than Norway, Switzerland or Netherlands, nations who came closest \cite{countrycomp} in terms of healthcare cost per capita. However, the huge expenditure does not necessarily buy Americans a better healthcare: in 2010, the number of practicing physicians per 1,000 people was only 2.4; whereas the corresponding number for the OECD countries was 3.1; there were 2.6 hospital beds per 1,000 people in 2009, whereas the OECD average was 3.4. In terms of life expectancy, the US ranked $28^{th}$ among the OECD countries in 2008, and it ranked below the average \cite{lifeex}. Moreover, a lot of the huge expense in the US is being attributed to unnecessary and inefficient measures, which include, but are not limited to: redundant medical tests \cite{seattle}, high cost of patented prescription drugs, infections caught from hospitals and frequent readmissions to hospitals. In 2011, the unnecessary expenditures added up to \$476 billion (18\%) to \$992 billion (37\%) of total (2.6 trillion) \cite{unnecessary}.\\

Policymakers and practitioners have started taking initiatives to reduce the overall expenditure and to tame its growth. Some of the measures suggested (and taken) are to increase the use of generic drugs as opposed to patented brand-name drugs, to identify services that were once considered good healthcare but now are suspected to lack evidence of benefit, to reduce the time spent due to administrative complexities, etc. Awareness about the usage of health data to derive useful insights for the patients, as well as for the service providers, is also on rise. As part of this initiative, the    
Department of Health and Human Services (HHS) took the Health Data Initiative \cite{hdi}, as a result of which, many health-related datasets, some of which are available from the insurance providers, are now publicly available on sites like \cite{healthdata}.\\

We attempt to address the problem of rise in healthcare expenditure by asking the following question: what factors increase an individual's expenses on healthcare? Is it because people develop certain chronic conditions? Is it because of age? Is it due to side-effects of drugs they have been prescribed? More specifically, given an individual's demographic and medical information, how well can we predict whether or not the individual's healthcare expense will rise next year? Although we formulated the problem as a binary classification problem, and experimented with various machine learning techniques to improve the classification accuracy, our goal was to go beyond simply achieving a high predictive accuracy: to identify how important the various factors are in increasing a person's healthcare expenditure.\\

Besides helping an individual, this can also be of interest to the patients/beneficiaries, the insurance providers and the healthcare service providers in ways we explain below:
\begin{itemize}
\item {\bf Beneficiaries:} If the beneficiaries know in advance that they are under high risk of an increase in expenditure for the next year, they can choose the insurance plans with higher deductible with more confidence. That way, although the annual out-of-pocket expense until the deductible is met would be higher, the beneficiary can benefit from a low monthly premium, and has to pay only 20-30\% co-insurance out-of-pocket once the deductible is met. 
\item {\bf Insurance providers:} The insurance providers can project the cost and revenue for the next year more accurately, given the data about the beneficiaries registered with them.
\item {\bf Healthcare service providers:} Hospital admissions and stays are intrinsically expensive. If doctors and hospitals know which patients are the high-risk ones, they can avoid readmission by taking preventive measures, e.g., arranging more frequent check-ups as outpatients, making the patients wear inexpensive sensors when they are discharged to monitor their conditions, etc.
\end{itemize}

{\bf Dataset: } We worked on the dataset ``CMS 2008-2010 Data Entrepreneurs$^\prime$ Synthetic Public Use File (DE-SynPUF)'' available at \cite{desynpuf}. This is an anonymized dataset obtained from Medicare \cite{medicare}, and has data about 6.87 million beneficiaries, their inpatient claims (claims filed on hospitalization), outpatient claims and prescription drug events. Since inpatient claims often form a significant fraction of an individual's annual healthcare expenses, we analyzed the expenditure on the inpatient claims. However, we use the other datasets (as well as some auxiliary datasets) to derive features based on the prescription drugs taken by the beneficiaries, the medical conditions they got diagnosed with as inpatients or even as outpatients, and the chronic conditions they had. One problem of healthcare datasets, in general, is that different aspects of a person's health are often siloed in different databases. This dataset is free from that problem, as we demonstrate with the following example taken from the data:\\
 
Beneficiary with ID \texttt{FFD1B248B8AAE4E5} had an inpatient cost of \$0 in 2008, which went up to \$5068 in 2009. The beneficiary data tells us that she developed three chronic conditions in 2009: Alzheimer's disease, heart failure and osteoporosis. The outpatient claims data tells us that she was diagnosed with the following conditions: stiffness of joint (ICD9 code 719.54), effusion of joint in the shoulder region (719.01), Tachypnea (rapid breathing, 786.06), hypertension (401.9) and Basilar artery syndrome (435.0) - the first two are probably related to osteoporosis. The prescription drug events data tell us that the beneficiary was prescribed with rivastigmine (used to treat mild to moderate dementia of the Alzheimer's type), warfarin sodium (anticoagulant, to prevent blood clotting) and prazosin hydrochloride (used to treat of hypertension): which shows that the drug data is indeed related to the data about the conditions diagnosed, and being diagnosed with so many conditions justify the significant increase in inpatient expenditure.\\

The dataset is anonymized to protect the privacy of the beneficiaries, but the various parts of the dataset are linked through beneficiary IDs that were created by applying a hash function on the original beneficiary IDs. We present more details of the dataset in Section~\ref{sec:data}.\\

{\bf Contributions:} Our contribution can be summarized as follows:
\begin{itemize}
\item We used a large, publicly available dataset with more than 114,000 beneficiaries and their health history for a span of three years (2008-2010) to investigate which ones, among more than 12,400 features, led to increase in expenditure as inpatient between 2008 and 2009. We formulated it as a binary classification problem. 
\item We experimented with various classification algorithms and finally chose six (gradient boosting machine \cite{Friedman01}, conditional inference tree \cite{HHZ06}, neural networks \cite{Ripley96}, SVM \cite{CV95}, logistic regression and naive Bayes) that performed best on a held-out test data set, and finally applied stacked generalization \cite{Wolpert92} as the ensemble technique. This achieved a sensitivity (recall) of 80\%, an overall accuracy of 77.56\% and a precision of 76.46\% on the test data set. This way, we showed that it is indeed possible to develop learning models which will predict with great degree of accuracy about whether an individual is going to incur higher or lower healthcare expenditure based on normally collected information.
\item Lastly, we identified major factors which are crucial in determining whether an individual is going to incur higher healthcare expenditure going forward. 
\end{itemize}

 
